### 2016-03-07 paper/icfp16-submission/semantics.tex
fixed genWitness

The witness generation procedure @genWitness@ is formalized in
Figure~\ref{fig:algo-gen-witness}.
%
Next, we describe its input and output, and how it
addresses the above challenges to search the space of possible
executions for general type error witnesses.

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound $k$ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
% Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
% denoting the list of \emph{top-level definitions}:
% %
% \begin{code}
%   let f1 = e1 in
%   let f2 = e2 in
%   ...
%   let fn = en in
% \end{code}
% %
% The witness generation procedure returns as output a list of (general)
% witness expressions $e$ each of which is of the form @fn v1 ... vn@.
Second, the target expression $e$ that contains the type error, and may be a curried function of multiple arguments.
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $k$ traces.
%
% In what follows, we refer to @fn@ as the \emph{target} of witness generation.

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\begin{mcode}
  eval :: $\triple{e}{\vsu}{\tsu}$ -> [$\triple{v \cup \stuck}{\vsu}{\tsu}$]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call
\hbox{|eval$\triple{e}{\vsu}{\tsu}$|} returns the \emph{list} of possible results
of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$ such that
$\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.
%
% We determine whether a given value is stuck by using the predicate:
% %
% \begin{mcode}
%   isStuck :: $v$ -> Bool
% \end{mcode}

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that
\hbox{|genArgs $e$|} yields a list of holes |[$v_1$,...,$v_n$]|
such that $\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes to the
target application until @eval@uating the term yields a
non-lambda value.
%
% The helper functions @witness@ and @close@ are used to respectively
% apply the parameters to the target and close the result under the top-level
% let-binders, prior to invoking \hbox{@eval@.}
%
% That is, @mkApps@ creates a nested sequence of applications in
% the usual left-associative style, and @mkLets@ takes a list of
% binders and a body expression, and creates a sequence of nested
% let-binders that close the body expression.

@@

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure @genWitness@,
which takes as input a bound @k@ and the target expression $e$, and
returns a list of witness expressions $\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target expression.
%
% @(2)@ we obtain a closed expression $e$ by closing the target with the
% computed holes.
%
@(2)@ we take the first @k@ traces returned by @eval@ on the target $\eapp{e}{vs}$
and
@(3)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following corollary of Theorem~\ref{thm:soundness}:

@@

\begin{figure}[t]
\centering
% $$
% \begin{array}{lcll}
% \genWitness{n}{e} &\defeq& \{ \resolve{e'}{\vsu} \spmid \triple{e'}{\vsu}{\tsu} \in \doeval{\genArgs{e}} \} & \\
% \end{array}
% $$

\begin{mcode}
genWitness :: Nat -> $e$ -> [$e$]
genWitness n $e$ = [$\resolve{\eapp{e}{vs}}{\vsu}$ | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs $e$                  -- (1)
   res = take n (eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$) $\ $  -- (2)
   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res] $\ \ $   -- (3)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}

>>>>

The witness generation procedure $\genWitnessN$ is formalized in
Figure~\ref{fig:algo-gen-witness}.
%
Next, we describe its input and output, and how it
addresses the above challenges to search the space of possible
executions for general type error witnesses.

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound $n$ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, the target expression $e$ that contains the type error,
and may be a curried function of multiple arguments.
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.

\paragraph{Modeling Semantics}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
$$
\evalN :: \triple{e}{\vsu}{\tsu} \rightarrow [\triple{v \cup \stuck}{\vsu}{\tsu}]
$$
%
Due to the non-determinism introduced by \gensym, a call
$\evalfn{\triple{e}{\vsu}{\tsu}}$ returns the \emph{list}
of possible results of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$
such that $\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure
$\genArgs{e}$ that yields a list of holes $[v_1, \ldots, v_k]$
such that $\eapp{e}{v_1 \ldots v_k}$ \emph{does not} evaluate to
a lambda.
%
This is achieved via a loop that keeps adding holes to the
target application until evaluating the term yields a
non-lambda value.
%
% The helper functions @witness@ and @close@ are used to respectively
% apply the parameters to the target and close the result under the top-level
% let-binders, prior to invoking \hbox{@eval@.}
%
% That is, @mkApps@ creates a nested sequence of applications in
% the usual left-associative style, and @mkLets@ takes a list of
% binders and a body expression, and creates a sequence of nested
% let-binders that close the body expression.

@@

\paragraph{Generating Witnesses}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure
$\genWitness{k}{e}$, which takes as input a bound $k$ and the
target expression $e$, and returns a list of witness expressions
$\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input program
gets stuck.
%
The search proceeds as follows.
%
\begin{enumerate}
  \item We invoke $\genArgsN$ to determine the number of
        parameters required by the (curried) target expression.

  \item We take the first $k$ traces returned by $\evalN$
        on the target $\eapp{e}{vs}$, and

  \item We extract the substitutions corresponding to the
        $\stuck$ traces, and use them to return the list
        of witnesses.
\end{enumerate}
%
We obtain the following corollary of Theorem~\ref{thm:soundness}:

@@

\begin{figure}[t]
\centering
$$
\begin{array}{lclr}
\genWitnessN       & :: & (\mathsf{Nat} \times e) \rightarrow 2^{e} & \\
\genWitness{n}{e}  & =  & \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \} & \\
\quad \mbox{\textbf{where}} &    & & \\
\quad \quad vs     & =  & \genArgs{e} & (1) \\
\quad \quad res    & =  & \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}} & (2) \\
\quad \quad \Sigma & =  & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)
\end{array}
$$
%% \begin{mcode}
%% $\genWitnessN :: (\mathsf{Nat} \times e) \rightarrow 2^{e}$
%% $\genWitness{n}{e} = \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \}$
  %% where
   %% $vs     = \genArgs{e}$               -- (1)
   %% $res    = \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}}$  -- (2)
   %% $\Sigma = \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\}$   -- (3)
%% \end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}
